---
title: "An Activity Classification Scheme Using The Random Forest Algorithm"
author: "Niyas Madappattu Alikutty"
date: "27 September 2015"
output: html_document
---
A classification scheme is developed using the Random Forest algorithm of R to predict the Class of activity. 

## Loading and Cleaning Data 
Loading the Data obatined as Data and summary of the data was examined.   
```{r}
Data<-read.csv("pml-training.csv", header=T, sep="," )
#unique(Data$num_window)
Data<-Data[order(Data$num_window),]
#summary(Data)
```

The Summary of the data indicates columns with more than 90% of the data missing from some columns. On closer examination of these columns and data indicates these are maximum, minimum, standard deviation, etc for each num_window. This data is also incomplete. 

As more than 90% percent of the data is missing, these colloumns are dropped. These data can easil be recreated from the other columns of the data. 
```{r}
#Removing columns with NAs and black spaces, these are columns which are mean, max, var etc. 
# these are columns that can be recreated from the Data3 
Data[Data==" "] <- NA
Data<-Data[, colSums(is.na(Data))<nrow(Data)*.7]
Data<-Data[, colSums(Data=="")<nrow(Data)*.7]
#summary(Data)
Data3<-Data[,-c(1:7)]
names(Data3)
```

### Training and Testing Data set

A training and testing data set is created with 60% of the data randomly selected from the Data3, the rest are used for testing. 
```{r}
#creating training and testing data set...
set.seed(1)
train=sample(1:nrow(Data3), nrow(Data3)*.6)
training<-Data3[train,]
testing<-Data3[-train,]
```

#Exploring the data 
```{r}
library(ggplot2)
p=ggplot(training)
p=p+geom_density(aes(x=roll_belt,fill=classe), alpha=0.1)
p=p+xlab('roll_belt')
p=p+ylab('Density')
p=p+ggtitle('Class Prediction')
p
```

While expoloring the data using various plots using predictors, it clear any other classification scheme will be difficult due to large number of predictors involved (52). So, the Random Forest algoirthm is used for this classification.Here, doParallel is used to multiple core available in the system. 

### Random Forest in Parallel 
doParallel package is used run randomForest in parallel process. The Random Forest algorithm are ideal for handling data with high dimensions. The Number of trees created are 900.  
```{r}
library(doParallel)# Parallelising using doParallel package for any 
library(randomForest)
cl <- makeCluster(3)# Using 3 Cores
registerDoParallel(cl)
set.seed(1)
x<-training[-ncol(training)]
y<-training$classe
rf.exercise <- foreach(ntree=rep(150,6), .combine=randomForest::combine,.packages='randomForest') %dopar% {
  randomForest(x,y, ntree=ntree, importance = T)
}
stopCluster(cl)
print(rf.exercise)
```

### Variable importance and Accuracy Measurment

```{r}
 rf.imp<-importance(rf.exercise)
 rf.imp[order(rf.imp[,6], decreasing=T),][1:15,]#15 predictors with importance
```

The confusion matrix is used to obtain the accuracy of the prediction. The predict function is used to make prediction on the testing dataset. The accuracy using the confusion matrix gives more than 99% accuracy and 95% confidence interval is within 99% accuracy. 
```{r cache=TRUE}
#confusion matrix
library(MASS)
library(caret)
 #testing data set
Predicttrain<-predict(rf.exercise, newdata=training)
confusionMatrix(Predicttrain, training$classe)

#testing data set
Prediction<-predict(rf.exercise, newdata=testing)
confusionMatrix(Prediction, testing$classe)
varImpPlot(rf.exercise, n.var=12, bg="red", main="Variable Importance")
```

### Testing with external data set
```{r}
Data_finaltest<-read.csv("pml-testing.csv", header=T, sep="," )
#Data_finaltest<-Data_finaltest[order(Data_finaltest$num_window),]
#Removing columns with NAs and black spaces, these are columns which are mean, max, var etc. 
# these are columns that can be recreated from the Data3 
Data_finaltest[Data_finaltest==" "] <- NA
Data_finaltest<-Data_finaltest[, colSums(is.na(Data_finaltest))<nrow(Data_finaltest)*.7]
Data_finaltest<-Data_finaltest[, colSums(Data_finaltest=="")<nrow(Data_finaltest)*.7]
#summary(Data)
Data3_finaltest<-Data_finaltest[,-c(1:7)]
Prediction<-predict(rf.exercise, newdata=Data3_finaltest)
Prediction
```

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(Prediction)
```


### Conclusion

A classification scheme was developed to predict the class of activity the participant is performing. The classification scheme predicts the activity on the testing set with more than 99% accurarcy. 
